Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-09 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Update wg_appendix_funding_spec_expert.tex

he power of the label? Getting the word out about RSE support using the label.

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.  And this brings up the flip side question, “Do scientists really want to be writing code?”
Career Path.
We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but tenure is always problematic for people of this class.

Skills and resources may vary between teams. To help resolve this, maintaining
high levels of communication between groups will be valuable. In the UK, there
are plans to permit resource sharing between institutional RSE groups.

Collaborative funding can be important for RSE groups, to ensure that research
leadership remains with the domain scientists. At NCAR, University partnerships
are required for submission of proposals, so collaboration is an essential part
of grant submission, and this will tend to bring developers and
scientists together. The UCL group also follows this approach, with all bids
requiring an academic collaborator.
Domain scientists and developers are funded together in a single proposal.
Another example of a success is the development of semantics and linked data in
support of Ocean Sciences. An EarthCube funded project pairs domain scientists
with RSEs and has been successful as the semantics attached have increased data
use and discovery significantly.

An alternative approach has been the provision of programming expertise as
part of national compute services. The US XSEDE project's extended collaboration
 service (ECS) is a set of developers who are paid with XSEDE funding,
 and are on “permanent” staff.  When PIs request allocations on XSEDE resources,
 there is a finite pool of developer time that can be awarded,
typically for one year only, and at partial effort, typically 20 percent or so.
The finite time allowed provides motivation for the scientist and their group to
work closely the developer and to become educated in what the developer is doing,
so they can sustain the effort, once the ECS period is over. This funding
mechanism can be highly efficient for scientific problems, because the developer
pool assembled by the research providers are, by definition, expert in the
characteristics of their specific resource, and can very quickly assess the
scientists needs, and what it will take to implement software that meets the
user’s needs. However, it does not develop capacity within institutions,
and indeed can have a ``chilling effect''.

@@

We would like to see these models applied outside High Performance Computing;
most scientific software is not destined to run on national cyberinfrastructure,
but needs similar support. The argument regarding making better use of expensive
hardware through software improvements has been useful politically, (and many
RSE groups are cited in organizations which host clusters for this reason), but
the time has come to make the case that software itself is a critical
cyberinfrastructure, and, with a much longer shelf-life than hardware, is itself
a capital investment.

@@

We will seek to identify and approach existing research programming organizations,
to get their permission to list them on a list of research software groups.
We will also look for examples of groups which have successfully become self-
sustaining following

@@

Financial support for an initial conference to bring together research software
groups to form an organization and create a resource sharing structure would
help to further this campaign.

In the longer term, funding organizations, especially non-governmental organizations
with the capability to effect innovation through seed funding, could provide
support to nucleate the creation of research software groups.

>>>>

The power of the label? Getting the word out about RSE support using the label.

Will research science developers be required, in the long run?
One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed. And this brings up the flip side question, “Do scientists really want to be writing code?”

Career Path.
We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but in the past tenure has often been problematic for people of this class.

Skills and resources may vary between teams. To help resolve this, maintaining
high levels of communication between groups will be valuable. In the UK, there
are plans to permit resource sharing between institutional RSE groups. Perhaps 
there are circumstances under which an RSE skill exchange could be arranged, either
formally or informally.

Collaborative funding can be important for RSE groups, to ensure that research
leadership remains with the domain scientists. At NCAR, University partnerships
are required for submission of proposals, so collaboration is an essential part
of grant submission, and this will tend to bring developers and
scientists together. The UCL group also follows this approach, with all bids
requiring an academic collaborator.

Domain scientists and developers are funded together in a single proposal.
Another example of a success is the development of semantics and linked data in
support of Ocean Sciences. An EarthCube-funded project pairs domain scientists
with RSEs and has been successful; the semantics attached have increased data
use and discovery significantly.

An alternative approach has been the provision of programming expertise as
part of national compute services. The US XSEDE project's extended collaboration
service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
“permanent” staff.  When PIs request allocations on XSEDE resources, there is a 
finite pool of developer time that can be awarded, typically for one year only, 
and at partial effort, typically 20 percent or so. The finite time allowed provides
motivation for the scientist and their group to work closely the developer and to 
become educated in what the developer is doing, so they can sustain the effort, once 
the ECSS period is over. This funding mechanism can be highly efficient for scientific
problems, because the developer pool assembled by the research providers are, by definition, 
expert in the characteristics of their specific resource, and can very quickly assess the
scientists needs, and what it will take to implement software that meets the user’s needs. 
However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
program, it should not be relied upon as a long term solution to acquiring this type of capacity.

@@

We would like to see these models applied outside High Performance Computing.
Most scientific software is not destined to run on national cyberinfrastructure,
but needs similar support. The argument regarding making better use of expensive
hardware through software improvements has been useful politically, (and many
RSE groups are cited in organizations which host clusters for this reason), but
the time has come to make the case that software itself is a critical
cyberinfrastructure, and, with a much longer shelf-life than hardware, is itself
a capital investment.

@@

We will seek to identify and approach existing research programming organizations,
to get their permission to list them on a list of research software groups.
Casual conversation during the meeting made it clear that although the title is not
widely used in the US, the this position is not rare. We spoke with several individuals
who, at distinct Universities, had RSEs (in effect if not in name) and these were funded
and under differing models.

We will also look for examples of groups which have successfully become self-
sustaining following.

In this respect, information gatheirng via a survey subsequent analysis could be very useful.
We would need to assemble a list of targetted individuals (what position and rank is
likely to know and care enough to respond?) Perhaps the Science Gateway Insitute has already acquire
information that could be helpful to advance this issue, and/or craft the proper survey and target
individuals.

@@

Financial support for an initial conference to bring together research software
groups to form an organization and create a resource sharing structure would
help to further this campaign. Funding to conduct and analyze a survey could also be quite useful
as knowing where we stand today, and what models are in use could fuel the ideas for further
development of developers in this category.

In the longer term, funding organizations, especially non-governmental organizations
with the capability to effect innovation through seed funding, could provide
support to nucleate the creation of research software groups. As noted above, Sloan has already
initiated one such program, and collaboration with Slaon or at least study of their methods and 
success or failure could be extremely useful in approaching Universities and other insitutions 
in funding this development track. It seems clear that if the value proposition can be made to
University adminstrators, this track could flourish with buy-in at the adminstrative level.
